PoPoolation Pipeline

### 2022-01-14 poolseq.Rmd
typos + formatting

## Filter Raw Sequences
This script uses the wrapper TrimGalore to remove Adapter/Barcode sequences and removes poor quality reads. 
This example dataset uses paired end sequencing so I used the flag (--paired) in TrimGalore. Because these example files had high coverage, I was able to use conservative filters. 
For example, the default length of reads is 20 bp (so if you read is less than 20 bp then the read will be removed). Since I had high coverage and long read lengths (150 bp), I used a minimum read length of 40 bp. 
The --max_n flag specifies how many "Ns" can be in the read before the read if discarded as "poor quality". Again, depending on your sequencing depth you may want to be more or less conservative. There is no default value specified in TrimGalore, but I used a highly conservative value of 1 (ie any instances of N in the read and that read will be discarded). This is probably more conservative than you need to be, because the read may still contain informative sequencing data.
Please note that these parameters are not specific to a Pool Seq dataset.
```{bash eval = FALSE}
#!/bin/bash

@@

## Compiling SNPs for all samples
This script compiles all the identified SNPs for each sample into a matrix. This sync file format is required for downstream analyses such as Fst, creating Manhattan plots, or CMH tests to be used in both the PoPoolation and PoPoolation2 pipelines. You will want to make sure that the fastq-type flag is set to sanger (this is for Phred quality score matching--yes, this is correct even if you used Illumina sequencing). The min-qual flag must be above 0, the default is 20 which gives you a 99% base call accuracy rate).

What is an Mpileup file?
Pileup format is a text-based format for summarizing the base calls of aligned reads to a reference sequence. This format facilitates visual display of SNP/indel calling and alignment. (Source: Wikipedia)
```{bash eval = FALSE}
#!/bin/bash

@@

You will want to play around with the min-coverage, max-coverage, and min-count based on the mean coverage of your samples/populations. Minimum coverages that are too high and max coverages that are too low may result in an empty output file, which is why its a good idea to have an estimate of your mean/min/max coverage for each sample/population. The minimum count must be less than the minimum coverage. You may want to modify this flag based on the number of individuals pooled per sample/population.

***Important considerations:***
--suppress-noninformative: Output files will not report results for windows with no SNPs or insufficient coverage. Including this flag will make it easier for downstream manipulation of the output file, such as creating figures.

@@

 '500' a maximum coverage of 500 will be used for all samples/populations
 '300,400,500' a maximum coverage of 300 will be used for the first sample/population, a maximum coverage of 400 for the second sample/population and so on
 '2%' the 2% highest coverages will be ignored, this value is independently estimated for every sample/population

***NOTE ABOUT MAX COVERAGE***
When genotyping individually barcoded samples, max coverage requirements are generally not recommended. However, pool-seq analyses require a max coverage for calculating Fst, cmh, allele frequency changes, etc. because allele frequency estimates are based on the number allele count, which are prone to PCR duplicates. PCR duplicates from a single individual in the pool may skew allele frequency estimates.

@@

## Calculate Fisher's Exact Test
This script will calculate significant allele frequency differences for each SNP position between two samples. Results can be plotted in a Manhattan plot. 
Depending on the experimental method, the --window-summary-method flag can be used to determine pvalue for windows of SNPs. The default is to multiply the pvalue of all SNPs for the window. Alternatively, you can specify how pvalues within a window should be summarized. Options include multiple, geometric mean, or median.  The summary of p-values within a window is dependent on the experimental design and how recent selection was expected to occur. For example, the default parameter might be appropriate if your treatment comparisons are geographically separated populations, where selection is expected to have occurred over hundreds of thousands of years. However, this might be inappropriate for short-term experimental evolution studies, such as the example dataset here. In this dataset, copepods were selected in the lab for 15-20 generations, therefore, we expect relatively large linkage blocks that prevent us from identifying the true targets of selection, and therefore I use the geometric mean option.

@@

Due to constraints of the test, you can only compare allele frequency changes for individual SNPs and not windows. In my data, I expected high levels of linkage disequilibrium in my samples because individuals were exposed to strong selection pressure. Depending on the amount of linkage disequilibrium you expect in your populations, you may want to account for the non-independence of SNPs following the CMH test. One way to do this, would be to calculate the mean (or geometric mean) p-value for all SNPs in a window. I provide some example R code below for how to do this.
I used the program SeqMonk to map cmh results to the genome and an annotated genome file I created with probes for the desired window size (i.e. 10,000 bp). Output files from SeqMonk will tell you which 10,000 bp window a particular SNP falls into. Thus, SNPs can be grouped for each 10,000bp window on each chromosome.
```{r eval=FALSE}
library("psych")

@@

cmh_p <-read.table("Tig.cmh", header=TRUE) #read in the original output from the CMH test, because SeqMonk output discards the important pvalue column.
colnames(cmh_p) <- c("CHR", "BP", "Allele", "1S", "2S", "3S", "4S", "5S", "U", "p") #rename columns based on sample names
cmh_p$ID <- paste(cmh_p$CHR,cmh_p$BP,sep=".")
cmh_p <- cmh_p[c(10,11)]

>>>>

## Filter Raw Sequences
The script below uses the wrapper TrimGalore to remove Adapter/Barcode sequences and removes poor quality reads. 
This example dataset uses paired end sequencing so I used the flag (--paired) in TrimGalore. Because these example files had high coverage, I was able to use conservative filters. 
For example, the default length of reads is 20 bp (so if a read is less than 20 bp then the read will be removed). Since I had high coverage and long read lengths (150 bp), I used a minimum read length of 40 bp. 
The --max_n flag specifies how many "Ns" can be in the read before the read if discarded as "poor quality". Again, depending on your sequencing depth you may want to be more or less conservative. There is no default value specified in TrimGalore, but I used a highly conservative value of 1 (ie any instances of N in the read and that read will be discarded). This is probably more conservative than you need to be, because the read may still contain informative sequencing data.
Please note that these parameters are not specific to a Pool Seq dataset.
```{bash eval = FALSE}
#!/bin/bash

@@

## Compiling SNPs for all samples
This script compiles all the identified SNPs for each sample into a matrix. This sync file format is required for downstream analyses such as Fst, creating Manhattan plots, or CMH tests to be used in both the PoPoolation and PoPoolation2 pipelines. You will want to make sure that the --fastq-type flag is set to "sanger" (this is for Phred quality score matching--yes, this is correct even if you used Illumina sequencing). The min-qual flag must be above 0, the default is 20 which gives you a 99% base call accuracy rate).

What is an Mpileup file?\
Pileup format is a text-based format for summarizing the base calls of aligned reads to a reference sequence. This format facilitates visual display of SNP/indel calling and alignment. (Source: Wikipedia)
```{bash eval = FALSE}
#!/bin/bash

@@

You will want to play around with the min-coverage, max-coverage, and min-count based on the mean coverage of your samples/populations. Minimum coverages that are too high and max coverages that are too low may result in an empty output file, which is why it's a good idea to have an estimate of your mean/min/max coverage for each sample/population. The minimum count must be less than the minimum coverage. You may want to modify this flag based on the number of individuals pooled per sample/population.

***Important considerations:***\
--suppress-noninformative: Output files will not report results for windows with no SNPs or insufficient coverage. Including this flag will make it easier for downstream manipulation of the output file, such as creating figures.

@@

 '500' a maximum coverage of 500 will be used for all samples/populations\
 '300,400,500' a maximum coverage of 300 will be used for the first sample/population, a maximum coverage of 400 for the second sample/population and so on\
 '2%' the 2% highest coverages will be ignored, this value is independently estimated for every sample/population

***NOTE ABOUT MAX COVERAGE***\
When genotyping individually barcoded samples, max coverage requirements are generally not recommended. However, pool-seq analyses require a max coverage for calculating Fst, cmh, allele frequency changes, etc. because allele frequency estimates are based on the number allele count, which are prone to PCR duplicates. PCR duplicates from a single individual in the pool may skew allele frequency estimates.

@@

## Calculate Fisher's Exact Test
This script will calculate significant allele frequency differences for each SNP position between two samples. Results can be plotted in a Manhattan plot. 
Depending on the experimental method, the --window-summary-method flag can be used to determine p-value for windows of SNPs. The default is to multiply the p-value of all SNPs for the window. Alternatively, you can specify how p-values within a window should be summarized. Options include multiple, geometric mean, or median.  The summary of p-values within a window is dependent on the experimental design and how recent selection was expected to occur. For example, the default parameter might be appropriate if your treatment comparisons are geographically separated populations, where selection is expected to have occurred over hundreds of thousands of years. However, this might be inappropriate for short-term experimental evolution studies, such as the example dataset here. In this dataset, copepods were selected in the lab for 15-20 generations, therefore, we expect relatively large linkage blocks that prevent us from identifying the true targets of selection, and therefore I use the geometric mean option.

@@

Due to constraints of the test, you can only compare allele frequency changes for individual SNPs and not windows. I expected high levels of linkage disequilibrium in my samples because individuals were exposed to strong selection pressure. Depending on the amount of linkage disequilibrium you expect in your populations, you may want to account for the non-independence of SNPs following the CMH test. One way to do this, would be to calculate the mean (or geometric mean) p-value for all SNPs in a window. I provide some example R code below for how to do this.
I used the program SeqMonk to map cmh results to the genome and an annotated genome file I created with probes for the desired window size (i.e. 10,000 bp). Output files from SeqMonk will tell you which 10,000 bp window a particular SNP falls into. Thus, SNPs can be grouped for each 10,000bp window on each chromosome.
```{r eval=FALSE}
library("psych")

@@

cmh_p <-read.table("Tig.cmh", header=TRUE) #read in the original output from the CMH test, because SeqMonk output discards the important p-value column.
colnames(cmh_p) <- c("CHR", "BP", "Allele", "1S", "2S", "3S", "4S", "5S", "U", "p") #rename columns based on sample names
cmh_p$ID <- paste(cmh_p$CHR,cmh_p$BP,sep=".")
cmh_p <- cmh_p[c(10,11)]
